Reactive SINDy: Discovering governing reactions from concentration data

### 2018-11-30 docs/manuscript_jcp.tex
update

The method presented in this work can identify underlying complex
reaction networks from concentration time series by following the
law of parsimony, i.e., by inducing sparsity in the resulting reaction
network. This promotes the interpretability of the model and avoids
overfitting. We formulate the problem as data-driven identification
of a dynamical system, which renders the method consistent with and
an extension of the framework of sparse identification of nonlinear
dynamics (SINDy)~\cite{Brunton2015}. Specifically, the problem of
identifying a reaction network from time traces of reactant concentrations
can be solved by finding a linear combination from a library of nonlinear
ansatz functions that each corresponds to a reaction acting on a set
of reactants. With this formulation, the reaction rates can be determined
\emph{via} regression. Sparsity is induced by equipping the regression
algorithms with a sparsity inducing regularization. SINDy was investigated,
generalized, and applied in many different ways, e.g., including control~\cite{brunton2016sparse}
(SINDYc), in the context of partial differential equations~\cite{rudy2017data},
updating already existing models~\cite{quade2018sparse} (abrupt-SINDy),
and looking into convergence properties~\cite{zhang2018convergence}.

@@

\begin{equation}
\dot{x}_{s}=\sum_{i}\beta_{s,0}^{(i)}+\sum_{i}\beta_{s,1}^{(i)}x_{i}+\sum_{i,j}\beta_{s,2}^{(i,j)}x_{i}x_{j}+\ldots\label{eq:lma}
\end{equation}
where the $\beta_{s,k}^{(\ldots)}$-values are constants belonging
to the reactions of order $k$. These rate constants however can incorporate
several underlying reactions at once. For example, the two reactions
\begin{align}
s_{1}\xrightharpoonup{\xi_{1}}s_{2}\label{eq:example_reaction_1}\\
s_{1}\xrightharpoonup{\xi_{2}}s_{3}\label{eq:example_reaction_2}
\end{align}
both contribute to $\dot{x}_{1}=\beta_{1,1}^{(1)}x_{1}=-(\xi_{1}+\xi_{2})x_{1}$.
To disentangle (\ref{eq:lma}) into single reactions, we choose a
library of $R$ possible ansatz reactions that each represent a single
reaction: 
\begin{align}
\textbf{y}_{r}(\textbf{x}(t))=\begin{pmatrix}y_{r,1}(\textbf{x}(t))\\
\vdots\\
y_{r,S}(\textbf{x}(t))
\end{pmatrix},\quad r=1,\ldots,R.
\end{align}
With this ansatz, the reaction dynamics (\ref{eq:lma}) becomes a
set of linear equations with unknown parameters $\xi_{r}$ that represent
the sought macroscopic rate constants: 
\begin{align}
\dot{\textbf{x}}_{i}(t)=\sum_{r=1}^{R}y_{r,i}(\textbf{x}(t))\xi_{r},\quad i=1,\ldots,S,\label{method:the-system}
\end{align}
where $\xi_{r}$ are the to-be estimated macroscopic rate constants.
The two reactions in the previous example (\ref{eq:example_reaction_1}-\ref{eq:example_reaction_2})
would be modeled by the ansatz functions
\begin{align*}
y_{1}(\mathbf{x}) & =(-x_{1},x_{1},0)^{\top}\\
y_{2}(\mathbf{x}) & =(-x_{1},0,x_{1})^{\top},
\end{align*}
illustrating that the values of the coefficients $\xi_{1}$ and $\xi_{2}$
can be used to decide whether a single reaction is present and to
what degree.

Now suppose we have measured the concentration vector (\ref{eq:concentration_vector})
at $T$ time points $t_{1}<\cdots<t_{T}$. We represent these data
as a matrix 
\begin{align}
\textbf{X}=\left(\begin{array}{cccc}
\mathbf{x}(t_{1}) & \mathbf{x}(t_{2}) & \cdots & \mathbf{x}(t_{T})\end{array}\right)^{\top}\in\mathbb{R}^{T\times S}.
\end{align}
Given this matrix, a library $\Theta:\mathbb{R}^{T\times S}\to\mathbb{R}^{T\times S\times R},\;\mathbf{X}\mapsto\begin{pmatrix}\theta_{1}(\mathbf{X}) & \theta_{2}(\mathbf{X}) & \cdots & \theta_{R}(\mathbf{X})\end{pmatrix}$
of $R$ ansatz reactions can be proposed with corresponding reaction
functions 
\begin{align}
\theta_{r}(\mathbf{X})=\begin{pmatrix}\textbf{y}_{r}(\mathbf{X}_{1*})^{\top}\\
\vdots\\
\textbf{y}_{r}(\mathbf{X}_{T*})^{\top}
\end{pmatrix}\in\mathbb{R}^{T\times S},\quad r=1,\ldots,R,\label{method:the-reactions}
\end{align}
where $\textbf{X}_{i*}$ denotes the $i$-th row in $\textbf{X}$.
Applying the concentration trajectory to the library yields $\Theta(\textbf{X})\in\mathbb{R}^{T\times S\times R}$.

@@

\ifthisispreprint
% ignore
\else
\onecolumngrid
\fi

\begin{figure}
\centering \includegraphics[width=0.82\doublecolumnwidth]{figures_jcp/3.pdf}
\caption{\label{fig:case-1-sparsity-pattern}Estimated reaction rates in the
system described in Sec.~\ref{sec:case-1}. The y and x axes contain
reaction educts and products, respectively. A circle at position $(i,j)$
represents a reaction $i\rightharpoonup j$ whose rate has a linear
relation with the area of the circle. The black outlines denote the
reactions with which the system was generated and contain the respective
rate value. Red crosses denote reactions that were used as additional
ansatz reactions. Blue circles are estimated by LSQ and orange circles
depict rates which were obtained by solving the minimization problem
(\ref{method:minimizationproblem}). The latter rates are subject
to a cutoff $\kappa=0.22$ corresponding to the green circle's area
under which a sparse solution with the correct processes can be recovered.
If a certain rate was estimated in both cases, two wedges instead
of one circle are displayed.}
\end{figure}
\ifthisispreprint
% ignore
\else
\twocolumngrid
\fi

@@

The value of the cutoff was determined by comparing the magnitude
of estimated rates and finding a gap, see Fig.~\ref{fig:reaction-rates-sorted}.
The hyperparameter pair $(\alpha,\lambda)$ was obtained by a grid
search and evaluating the difference $\|\hat{\Xi}_{\alpha,\lambda}-\Xi\|_{1}$,
where $\hat{\Xi}_{\alpha,\lambda}$ is the estimated model under a
particular hyperparameter choice and $\Xi$ is the ground truth. If
the ground truth is unknown, a hyperparameter pair can be estimated
by utilizing cross-validation as in the following sections.

@@

The corresponding time series are depicted in Fig.~\ref{fig:case3-convergence}a.
The gray graph corresponds to a sample trajectory generated by the
Gillespie SSA. For both initial conditions the same time step of $\tau=3\cdot10^{-3}$
has been applied, amounting to $2\cdot667=1334$ frames. Once the
data matrices 
\[
\mathbf{X}_{1}=\begin{pmatrix}\boldsymbol{x}_{1}(t_{1}) & \cdots & \boldsymbol{x}_{1}(t_{667})\end{pmatrix},\;\mathbf{X}_{2}=\begin{pmatrix}\boldsymbol{x}_{2}(t_{1}) & \cdots & \boldsymbol{x}_{2}(t_{667})\end{pmatrix}
\]
 and the corresponding derivatives $\dot{\mathbf{X}}_{1}$, $\dot{\mathbf{X}}_{2}$
have been obtained, the frames are concatenated so that 
\[
\mathbf{X}=\begin{pmatrix}\boldsymbol{x}_{1}(t_{1}) & \boldsymbol{x}_{2}(t_{1}) & \cdots & \boldsymbol{x}_{1}(t_{667}) & \boldsymbol{x}_{2}(t_{667})\end{pmatrix},
\]
analogously for $\dot{\mathbf{X}}$.

@@

\ifthisispreprint
% ignore
\else
\onecolumngrid
\fi

\begin{table}[h!]

@@

\end{table}

\ifthisispreprint
% ignore
\else
\twocolumngrid
\fi

>>>>

The method presented in this work can identify underlying complex
reaction networks from concentration time series by following the
law of parsimony, i.e., by inducing sparsity in the resulting reaction
network. This promotes the interpretability of the model and avoids
overfitting. We formulate the problem as data-driven identification
of a dynamical system, which renders the method consistent with and
an extension of the framework of sparse identification of nonlinear
dynamics (SINDy)~\cite{Brunton2015}. Specifically, the problem of
identifying a reaction network from time traces of reactant concentrations
can be solved by finding a linear combination from a library of candidate
nonlinear functions (ansatz functions) that each corresponds to a reaction acting on a set
of reactants. With this formulation, the reaction rates can be determined
\emph{via} regression. Sparsity is induced by equipping the regression
algorithms with a sparsity inducing regularization. SINDy was investigated,
generalized, and applied in many different ways, e.g., including control~\cite{brunton2016sparse}
(SINDYc), in the context of partial differential equations~\cite{rudy2017data},
updating already existing models~\cite{quade2018sparse} (abrupt-SINDy),
and looking into convergence properties~\cite{zhang2018convergence}.

@@

\begin{equation}
\dot{x}_{s}=\sum_{i}\beta_{s,0}^{(i)}+\sum_{i}\beta_{s,1}^{(i)}x_{i}+\sum_{i,j}\beta_{s,2}^{(i,j)}x_{i}x_{j}+\ldots\label{eq:lma}
\end{equation}
where the $\beta_{s,k}^{(\ldots)}$-values are constants belonging
to the reactions of order $k$. These rate constants however can incorporate
several underlying reactions at once. For example, the two reactions
\begin{align}
s_{1}\xrightharpoonup{\xi_{1}}s_{2}\label{eq:example_reaction_1}\\
s_{1}\xrightharpoonup{\xi_{2}}s_{3}\label{eq:example_reaction_2}
\end{align}
both contribute to $\dot{x}_{1}=\beta_{1,1}^{(1)}x_{1}=-(\xi_{1}+\xi_{2})x_{1}$.
To disentangle (\ref{eq:lma}) into single reactions, we choose a
library of $R$ possible ansatz reactions that each represent a single
reaction: 
\begin{align}
\textbf{y}_{r}(\textbf{x}(t))=\begin{pmatrix}y_{r,1}(\textbf{x}(t))\\
\vdots\\
y_{r,S}(\textbf{x}(t))
\end{pmatrix},\quad r=1,\ldots,R.
\end{align}
With this ansatz, the reaction dynamics (\ref{eq:lma}) becomes a
set of linear equations with unknown parameters $\xi_{r}$ that represent
the sought macroscopic rate constants: 
\begin{align}
\dot{\textbf{x}}_{i}(t)=\sum_{r=1}^{R}y_{r,i}(\textbf{x}(t))\xi_{r},\quad i=1,\ldots,S,\label{method:the-system}
\end{align}
where $\xi_{r}$ are the to-be estimated macroscopic rate constants.
The two reactions in the previous example (\ref{eq:example_reaction_1}-\ref{eq:example_reaction_2})
would be modeled by the functions
\begin{align*}
y_{1}(\mathbf{x}) & =(-x_{1},x_{1},0)^{\top},\\
y_{2}(\mathbf{x}) & =(-x_{1},0,x_{1})^{\top},
\end{align*}
illustrating that the values of the coefficients $\xi_{1}$ and $\xi_{2}$
can be used to decide whether a single reaction is present and to
what degree.

Now suppose we have measured the concentration vector (\ref{eq:concentration_vector})
at $T$ time points $t_{1}<\cdots<t_{T}$. We represent these data
as a matrix 
\begin{align}
\textbf{X}=\left(\begin{array}{cccc}
\mathbf{x}(t_{1}) & \mathbf{x}(t_{2}) & \cdots & \mathbf{x}(t_{T})\end{array}\right)^{\top}\in\mathbb{R}^{T\times S}.
\end{align}
Given this matrix, a library $\Theta:\mathbb{R}^{T\times S}\to\mathbb{R}^{T\times S\times R},\;\mathbf{X}\mapsto\begin{pmatrix}\theta_{1}(\mathbf{X}) & \theta_{2}(\mathbf{X}) & \cdots & \theta_{R}(\mathbf{X})\end{pmatrix}$
of $R$ candidate (ansatz) reactions can be proposed with corresponding reaction
functions 
\begin{align}
\theta_{r}(\mathbf{X})=\begin{pmatrix}\textbf{y}_{r}(\mathbf{X}_{1*})^{\top}\\
\vdots\\
\textbf{y}_{r}(\mathbf{X}_{T*})^{\top}
\end{pmatrix}\in\mathbb{R}^{T\times S},\quad r=1,\ldots,R,\label{method:the-reactions}
\end{align}
where $\textbf{X}_{i*}$ denotes the $i$-th row in $\textbf{X}$.
Applying the concentration trajectory to the library yields $\Theta(\textbf{X})\in\mathbb{R}^{T\times S\times R}$.

@@

%\ifthisispreprint
% ignore
%\else
%\onecolumngrid
%\fi

\begin{figure*}
\centering \includegraphics[width=0.82\doublecolumnwidth]{figures_jcp/3.pdf}
\caption{\label{fig:case-1-sparsity-pattern}Estimated reaction rates in the
system described in Sec.~\ref{sec:case-1}. The y and x axes contain
reaction educts and products, respectively. A circle at position $(i,j)$
represents a reaction $i\rightharpoonup j$ whose rate has a linear
relation with the area of the circle. The black outlines denote the
reactions with which the system was generated and contain the respective
rate value. Red crosses denote reactions that were used as additional
ansatz reactions. Blue circles are estimated by LSQ and orange circles
depict rates which were obtained by solving the minimization problem
(\ref{method:minimizationproblem}). The latter rates are subject
to a cutoff $\kappa=0.22$ corresponding to the green circle's area
under which a sparse solution with the correct processes can be recovered.
If a certain rate was estimated in both cases, two wedges instead
of one circle are displayed.}
\end{figure*}
%\ifthisispreprint
% ignore
%\else
%\twocolumngrid
%\fi

@@

The value of the cutoff $\kappa$ was determined by comparing the magnitude
of estimated rates and finding a gap, see Fig.~\ref{fig:reaction-rates-sorted}.
The hyperparameter pair $(\alpha,\lambda)$ was obtained by a grid
search and evaluating the difference $\|\hat{\Xi}_{\alpha,\lambda}-\Xi\|_{1}$,
where $\hat{\Xi}_{\alpha,\lambda}$ is the estimated model under a
particular hyperparameter choice and $\Xi$ is the ground truth. If
the ground truth is unknown, a hyperparameter pair can be estimated
by utilizing cross-validation as in the following sections.

@@

The corresponding time series are depicted in Fig.~\ref{fig:case3-convergence}a.
The gray graph corresponds to a sample trajectory generated by the
Gillespie SSA. For both initial conditions the same time step of $\tau=3\cdot10^{-3}$
has been applied, amounting to $2\cdot667=1334$ frames. Once the
data matrices 
\[
\mathbf{X}_{1}=\begin{pmatrix}\boldsymbol{x}_{1}(t_{1}) & \cdots & \boldsymbol{x}_{1}(t_{667})\end{pmatrix},\;\mathbf{X}_{2}=\begin{pmatrix}\boldsymbol{x}_{2}(t_{1}) & \cdots & \boldsymbol{x}_{2}(t_{667})\end{pmatrix}
\]
 and the corresponding derivatives $\dot{\mathbf{X}}_{1}$, $\dot{\mathbf{X}}_{2}$
have been obtained, the frames are concatenated so that 
\[
\mathbf{X}=\begin{pmatrix}\boldsymbol{x}_{1}(t_{1}) & \cdots & \boldsymbol{x}_{1}(t_{667}) & \boldsymbol{x}_{2}(t_{1}) & \cdots & \boldsymbol{x}_{2}(t_{667})\end{pmatrix},
\]
analogously for $\dot{\mathbf{X}}$.

@@

%\ifthisispreprint
% ignore
%\else
%\onecolumngrid
%\fi

\begin{table*}[h!]

@@

\end{table*}

%\ifthisispreprint
% ignore
%\else
%\twocolumngrid
%\fi
